---
title: "Phyllostomus hastatus, Bocas del Toro: Simulation of commuting tracks using ctmm"
author: "Mariëlle"
format: 
  html:
    embed-resources: true
editor: source
---

## Session set-up

```{r}

library(ctmm)
library(ggplot2)
library(sf)
library(suntools)
library(plyr)
library(ggspatial)
library(geosphere)
library(lubridate)
library(momentuHMM)

proj.ll <- '+proj=longlat +datum=WGS84'

adm <- rbind(st_read('data/environmental/gadm41_PAN_shp/gadm41_PAN_0.shp', quiet=T),
             st_read('data/environmental/gadm41_CRI_shp/gadm41_CRI_0.shp', quiet=T))

coastline <- st_union(adm)

```

### Cave locations

```{r}

cave.loc <- data.frame(
  cave=c('lagruta', 'ajcave', 'muddycave'),
  long=c(-82.271541, -82.274955, -82.264753),
  lat=c(9.396448, 9.440312, 9.387233)
)

```

## Data preparation

Importing the data with annotated behavior, and I somehow kept a couple of unnecessary steps that Camila had already done in her own way:

```{r}

(load('data/tracking_data_annotated/Phyllostomus2021-2022_HMMbehaviors.RData'))

# rename caves
bats_behaviors$cave[bats_behaviors$cave=='LG'] <- 'lagruta'
bats_behaviors$cave[bats_behaviors$cave=='MT'] <- 'ajcave'

# remove columns that contain no data
bats_behaviors <- bats_behaviors[,apply(bats_behaviors, 2, function(x){any(!is.na(x))})]

# assign season (I'm currently assuming that December is rather wet than dry)
bats_behaviors$season <- ifelse(bats_behaviors$month %in% c('Aug', 'Dec'), 'wet', 'dry')

# assign "day" by calculating time of day relative to solar noon
p <- st_as_sf(bats_behaviors, coords=c('location_long', 'location_lat'), crs=st_crs(proj.ll))
noon <- solarnoon(p, bats_behaviors$timestamp, POSIXct.out=T)

# fractional hours after noon
bats_behaviors$tod <- as.numeric(difftime(bats_behaviors$timestamp, noon$time, units = 'h'))%%24

# split individual tracks into separate days
bats_behaviors <- ddply(bats_behaviors, 'individual_id', function(x){
  x$day <- c(0, cumsum(diff(x$tod)<0)) + 1
  return(x)
})

```

## Identify outbound and inbound commute

Camila had already applied a HMM to classify the behaviours of individuals at points using step length & turning angles. Here I took a slightly different approach that makes it easier to separate the outbound commute, away from the cave, from the commute back to the cave. I did this by replacing turning angle with angle to the colony, which takes values around $\pi$ when individuals fly away from the cave, and 0 when they fly towards it. Step length contains sufficient information to separate flight from other behaviour, so it was not necessary to do a lot of optimisation to get to a model that looks pretty good. I thought of making a couple of additions to the model, such as fixing the order of states or looking into effects of season & cave on direction of travel, but that turned out to be complicating things where not necessary.

```{r}
# fit HMM to time series of locations representing bat-days
bats_behaviors$ID <- bats_behaviors$ID_batday

# prepare cave locations for angle calculation
caves <- as.matrix(cave.loc[,2:3])
dimnames(caves)[[1]] <- cave.loc$cave

# convert data to momentuHMM format
bats <- prepData(bats_behaviors[,!names(bats_behaviors) %in% c('x', 'y', 'step', 'angle')], 
                 coordNames = c('location_long', 'location_lat'), 
                 covNames=c('cave'), type = 'LL', 
                 centers = caves)

# make new angle covariate choosing the angle to the respective cave for each individual
bats$cave.angle <- unlist(lapply(1:nrow(bats), function(j){
  keep <- ifelse(bats$cave[j]=='lagruta', bats$lagruta.angle[j], 
                 ifelse(bats$cave[j]=='ajcave', bats$ajcave.angle[j], bats$muddycave.angle[j]))
}))

# starting parameters
par <- getParDM(data=bats, 
                nbStates=3,
                dist=list(step='gamma', cave.angle='wrpcauchy'), 
                estAngleMean = list(cave.angle=T),
                Par=list(step=c(0.1,1.25,1.25,0.5,2,2,0.5,0.01,0.01),
                         cave.angle=c(pi-0.01, 0, pi-0.01, 0.3,0.7,0.7)),
                userBounds = list(cave.angle=matrix(c(rep(c(-pi,pi), 3), rep(c(0,1), 3)), ncol=2, byrow=T))) # DM=DM, 

# fit HMM
hmm2 <- fitHMM(data=bats,
              nbStates=3,
              dist=list(step='gamma', cave.angle='wrpcauchy'), 
              estAngleMean = list(cave.angle=T),
              Par0=par,
              stateNames=c('other', 'inbound commute', 'outbound commute'))

hmm2

# does the model look any good?
plot(hmm2, plotTracks=T, animals=sample(length(unique(bats$ID)), 12), ask=F)

```

It turns out bats are quite similar to (most) birds when it comes to a straightforward HMM into flight/non-flight behaviour, and I'm quite happy with how the inbound and outbound commute separate as well. In the next step, I will retain for each bat-day the data from the first to the last location considered "outbound commute"; i.e. this will contain some foraging locations as well. 

```{r}

bats$state <- viterbi(hmm2)

# which state is outbound commute?
ca <- hmm2$mle$cave.angle
(keep <- which(ca[2,]>0.7 & abs(round(ca[1,]))==3))

commute.outward <- ddply(bats, 'ID', function(x){
  if(any(x$state==3)){
    keep <- which(x$state==3)
    return(x[min(keep):max(keep),])
  }else{return(NULL)}
})

```

What remains?

```{r}
#| echo: false
#| waning: false

ggplot() +
  geom_sf(data=adm, fill='antiquewhite', colour='antiquewhite4', linewidth=0.2) +  xlim(range(commute.outward$x)) +
  ylim(range(commute.outward$y)) +
  theme_light() +
  theme(axis.title=element_blank(), 
        panel.grid=element_blank(),
        plot.background = element_rect(fill = "white", color = NA), 
        panel.background = element_rect(fill = '#dcebf0', color = NA), 
        legend.background = element_rect(fill = "white", color = NA), ##f5f5f2
        panel.border = element_blank()) +
  annotation_scale(location = "tr", line_width = 1) +
  labs(x='Longitude', y='Latitude') +
  geom_path(data=commute.outward, aes(x=x, y=y, group=ID, colour=cave)) +
  scale_colour_viridis_d(end=0.8, option='C', name='', direction=-1) +
  guides(colour=guide_legend(override.aes = list(linewidth=2)))

```

Now that we have pulled out the outbound commute, I can fit the IOU-ctmms to the data for the simulations. First, I will remove any outbound commute that starts on the mainland, just to make sure we don't have any stray return movements remaining in the data. 

```{r}
#| eval: false

# re-arrange data for use with ctmm
commute.outward$individual.local.identifier <- commute.outward$ID
commute.outward$location.long <- commute.outward$x
commute.outward$location.lat <- commute.outward$y

# check starting point
panama <- st_cast(adm[adm$COUNTRY=='Panama',], 'POLYGON')
panama$area <- st_area(panama)
mainland <- panama[which.max(panama$area),]
p <- st_as_sf(commute.outward, coords=c('location.long', 'location.lat'), crs=st_crs(mainland))
commute.outward$mainland <- st_within(p, mainland, sparse=F)[,1]

# remove bat-days that have outbound commute starting on the mainland
commute.outward <- ddply(commute.outward, 'ID', function(x){
  if(x$mainland[1]){return(NULL)}
  # if(nrow(x)<5){return(NULL)}
  return(x)
})

# convert to telemetry-object
tel.all <- as.telemetry(commute.outward)
tel.all <- tel.all[!is.na(summary(tel.all)[,1])]

# we need a couple of data points to fit a model, and we want to make sure that we base it on a bit of actual commute rather than foraging
# retain bat-days with at least 5 commuting locations
commute.prop <- ddply(commute.outward, 'ID', function(x){data.frame(n=sum(x$state==keep), nrow=nrow(x))})
tel.all <- tel.all[names(tel.all) %in% commute.prop$ID[commute.prop$n>5]]

# remove tracks with less than 10 data points
tel.all <- tel.all[sapply(tel.all, nrow)>10]

# this leaves 218 tracks to fit our model to!

ctmm.gps <- lapply(names(tel.all), function(id){
  # print(id)
  tel <- tel.all[[id]]
  var <- variogram(tel, dt=c(2%#%'minutes'))
  PROTO <- ctmm(range=FALSE)
  GUESS <- ctmm.guess(tel, var=var, interactive=FALSE, CTMM=PROTO)
  m <- try(ctmm.select(tel, CTMM=GUESS, cores=2, verbose=F))
  return(m)
}); names(ctmm.gps) <- sapply(ctmm.gps, function(x){x@info$identity})

# save models
save(tel.all, ctmm.gps, file='iou_models_outbound_commute.RData')

```

Some means from the models:

```{r}
#| echo: false

# replaced ballistic with iou just in the filename above
load('ballistic_models_outbound_commute.RData')

```
```{r}
#| fig-height: 8

meta(ctmm.gps, variable='speed')
meta(ctmm.gps, variable='diffusion')

```

## Simulations

Now everything should be ready for the simulations of alternative outbound commuting tracks, apart from one thing: we need to determine for how many steps we want to simulate from the model. One issue with the outbound commute is that it was not always fully captured if the GPS started up late. To avoid assuming to small a number of steps, use both the mean of the time spent on inbound and outbound commute to determine a empirical distribution of steps that we can sample from.

```{r}
# time in return commute (gets us the entire distance rather than just part of the outbound commute)

# which states are commuting states?
(commute.state <- which(ca[2,]>0.7))

commute.time <- timeInStates(hmm2, by='ID')
commute.time$track.dur <- unlist(lapply(commute.time$ID, function(id){
  diff(range(as.numeric(bats$timestamp[bats$ID==id])))
}))
commute.time <- (apply(commute.time[,commute.state+1], 1, sum)*commute.time$track.dur) / 2
commute.time <- round(commute.time[commute.time>0]/120)

```


```{r}
#| echo: false
#| warning: false

ggplot(data.frame(steps=commute.time), aes(x=steps)) + 
  geom_histogram(breaks=seq(0,130,5), fill='grey90', colour='grey20') +
  scale_x_continuous(breaks=seq(0,125,25)) +
  theme_light() + labs(x='Number of commuting steps')

ggsave(filename='empirical_step_dist.png', width=88, height=80, unit='mm')

```
```{r}

cs <- expand.grid(cave=c('lagruta', 'ajcave', 'muddycave'),
                  season=c('dry', 'wet'), KEEP.OUT.ATTRS=F)

check.n <- ddply(commute.outward, 'ID', nrow)
keep.id <- commute.outward[commute.outward$ctmm.gps %in% names(ctmm.gps),]

sim <- ddply(cs, c('cave', 'season'), function(x){
  cv <- x$cave; se <- x$season
  print(paste(cv, se, sep=' - '))
  ids <- unique(commute.outward$ID[commute.outward$cave==cv & commute.outward$season==se])
  ids <- as.character(ids[ids %in% names(ctmm.gps)])
  if(length(ids)==0){return(NULL)}
  
  rbind.fill(lapply(sample(ids, 5000, replace=T), function(id){
    steps <- sample(commute.time, 1)
    sim <- simulate(ctmm.gps[[id]], t=seq(0,steps*120,120))
    sim <- as.data.frame(sim@.Data)
    sim <- sim[nrow(sim),2:3]; names(sim) <- c('x', 'y')
    sim$n <- steps
    return(sim)
  }))

})

p <- st_as_sf(sim, coords=c('x', 'y'), crs=st_crs(tel.all[[1]]@info$projection))
p.ll <- st_transform(p, st_crs(proj.ll))
sim$long <- st_coordinates(p.ll)[,1]
sim$lat <- st_coordinates(p.ll)[,2]
sim$land <- st_within(p.ll, adm, sparse=F)[,1]

sim$cave.label <- factor(sim$cave, labels=c('La Gruta', 'Aj Cave', 'Muddy Cave'))

# end-of-commuting locations
# foraging <- bats_behaviors[bats_behaviors$behav2=='foraging',]
foraging <- ddply(bats[!is.na(bats$cave),], 'ID', function(x){
  if(any(x$state==3)){
    keep <- which(x$state==3)
    return(x[max(keep)+1,])
  }else{return(NULL)}
})
foraging <- foraging[complete.cases(foraging$ID),]

foraging$cave2 <- ifelse(foraging$cave=='lagruta', 'Colony 1', 
                         ifelse(foraging$cave=='ajcave', 'Colony 2', 'Colony 3'))
foraging$season <- ifelse(foraging$season=='dry', 'dry season', 'wet season')
foraging$group <- paste(foraging$cave2, foraging$season, sep='-')

p1 <- ggplot() +
  geom_sf(data=adm, fill='white', colour='black', linewidth=0.2) +
  theme_light() +
  theme(axis.title=element_blank(), 
        panel.grid=element_blank(),
        plot.background = element_rect(fill = "white", color = NA), 
        panel.background = element_rect(fill = 'grey90', color = NA), 
        legend.background = element_rect(fill = "white", color = NA), ##f5f5f2
        panel.border = element_blank()) +
  labs(x='Longitude', y='Latitude') +
  # geom_point(data=sim[sim$land,], aes(x=long, y=lat), colour='grey20', size=0.5) +
  geom_point(data=sim[sim$land,], aes(x=long, y=lat), size=0.05, alpha=1, show.legend=T) + 
  geom_density_2d(data=sim[sim$land,], aes(x=long, y=lat), show.legend=F, colour=viridisLite::inferno(9)[6], linewidth=0.2) +
  # scale_colour_manual(values=c(NA, viridisLite::viridis(7))) +
  geom_point(data=foraging, aes(x=x, y=y, colour=group), size=1, alpha=0.7) +
  guides(colour=guide_legend(override.aes = list(alpha=1, size=2, shape=16))) + 
  scale_colour_manual(name='', values=viridisLite::viridis(12)[c(2,4,7,9,12)]) +
  scale_x_continuous(expand=c(0,0), limits=c(min(sim$long[sim$land]), -82)) + 
  scale_y_continuous(expand=c(0,0), limits=c(9, max(sim$lat[sim$land]))) +
  annotation_scale(location = "tr", line_width = 1) #+
  # facet_grid(factor(season, labels=c('dry season', 'wet season'))~cave.label)

ggsave(plot=p1, filename='simulated_vs_real_foraging_locations.pdf', width=174, height=100, unit='mm', dpi=600)

foraging$angle <- unlist(lapply(1:nrow(foraging), function(j){
  bearing(cave.loc[cave.loc$cave==foraging$cave[j],2:3], foraging[j,c('x', 'y')])
}))

sim <- ddply(sim, 'cave', function(x){
  cave <- unique(x$cave)
  p <- st_as_sf(x, coords=c('long', 'lat'), crs=st_crs(proj.ll))
  x$angle <- bearing(cave.loc[cave.loc$cave==cave,2:3], st_coordinates(p))
  return(x)
})

sim$cave2 <- ifelse(sim$cave=='lagruta', 'Colony 1', 
                          ifelse(sim$cave=='ajcave', 'Colony 2', 'Colony 3'))
sim$season <- ifelse(sim$season=='dry', 'dry season', 'wet season')

library(viridisLite)

p2 <- ggplot(sim) +
  geom_histogram(aes(x=(angle+360)%%360), fill='grey90', colour='grey20', breaks=seq(0,360,10)) +
  facet_grid(season~cave2) +
  theme_light() +
  geom_vline(data=foraging, aes(xintercept=(angle+360)%%360, colour=season), linewidth=0.2, alpha=0.5) +
  scale_colour_manual(name='', values=c('dry season'=inferno(9)[7], 'wet season'=mako(9)[3])) +
  guides(colour=guide_legend(override.aes = list(linewidth=2, alpha=1))) +
  labs(x='Angle from cave (degrees)') +
  scale_y_continuous(expand=c(0.001,5)) +
  scale_x_continuous(breaks=seq(0,360,90)) +
  theme(legend.position='none', strip.background = element_rect(fill=NA),
        strip.text=element_text(colour='black'))

ggsave(plot=p2, filename='angles.pdf', width=174, height=80, unit='mm', dpi=600)

library(patchwork)

p1 + p2 + plot_layout(ncol=1) + plot_annotation(tag_levels='A')

ggsave(filename='simulation_plots.pdf', width=174, height=140, unit='mm')


```

